gazeHMM: Parameter recovery simulation

### 2020-05-08 simulation/preregistration_parameter_recovery.Rmd
update parameter recovery preregistration

In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM. The purpose of this algorithm is to classify eye movements into distinct events. These events include fixations, saccades, and optionally postsaccadic oscillations (PSOs) and smooth pursuits. The full algorithm will be available on GitHub (www.github.com/maltelueken/gazeHmm) upon publication.

@@

The HMM estimates non-observable states that have generated the data. In the context of eye-tracking, each state corresponds to an eye movement event. In gazeHMM, the model can have between two and four states. State one always corresponds to fixations, state two always to saccades, state three to PSOs, and state four to smooth pursuits. Moreover, the HMM is multivariate but conditionally independent on the states. Thus, it has three independent response models for every state describing velocity, acceleration, and angle. The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization). The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to describe angle. Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively). Note that no covariates are included in the model, only intercepts for the parameters are estimated.

@@

To estimate parameter recovery of the HMM, we repeatedly generate data with the model under different conditions and different numbers of states. For every condition and number of states $k \in \{2, 3, 4\}$, the model will generate $D = 100$ data sets with parameters varied in a specified interval in equidistant steps. The same model is then applied to estimate the parameters from the generated data. Then, the estimated parameters are compared with the "true" parameters that generated the data. We will assess parameter recovery descriptively by computing the mean squared error between the estimated and true parameter values separately for each parameter. Additionally, we will apply a bivariate linear regression for each parameter with the true parameter values as the independent and the estimated parameter values as the dependent variable. For a parameter to be recovered, we expect the regression intercepts to be close to zero and the regression slopes to be close to one. Next to parameter recovery, we will also examine the accuracy of the classficiation through the model. Therefore, we will compare the true states with the estimated states for each data set by using Cohen's Kappa. We will treat Kappa values above 0.8 as satisfactorily accurate.

The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilities. To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$. Mean parameters of the von-Mises distribution will always start at their true value.

In the first part of the simulation study, we will estimate parameter recovery under ideal conditions. These imply a medium sample size of $N = 2500$ samples per data set, no noise added to the data, and no missing samples. The parameters that we will use to generate the data are displayed in Table 1. We will vary one parameter at a time on the given interval while setting the other parameters to their fixed value. The initial state probabilities will always stay at their fixed value. We will only vary the transition probabilities for staying in the same state. The remaining probability mass will be evenly distributed across the probabilities for switching to a different state (see Table 1). For two states 10, for three states 15, and for four states 20 parameters will be varied one at a time. This simulation will result in $D_{total} = 100 \times (10+15+20) = 4500$ recoveries.

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. The model parameters will be set to the fixed values from Table 1. For sample sizes of $N \in \{500, 2500, 10000\}$, we will generate data sets and add white noise. To velocity and acceleration, we will add Gaussian noise with $\mu = 0$ and $\sigma \in [1,25]$. Noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ will be added to angle. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the third part, we will increase the variation in the starting values used for parameter estimation. The model parameters will be set to the fixed values from Table 1. For the shape, scale, and concentration parameters, we will increase the scale parameters of the gamma distributions to $\beta \in \{2, 4, 8\}$ all at the same time. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

In the last part, we will set intervals of the generated data to be missing. The model parameters will be set to the fixed values from Table 1. The size of the missing data interval will be $m \in [1,100]$ and the number of intervals will be $n_{miss} \in \{1, 3, 5\}$. The sample size will be $N = 2500$ and no noise will be added to the data. This simulation will result in $D_{total} = 100 \times 3 \times 3 = 900$ recoveries.

@@

>>>>

In this document, we are preregistering a simulation study to estimate the parameter recovery of a hidden Markov model (HMM) which is part of our recently developed algorithm gazeHMM. The purpose of this algorithm is to classify eye movements into distinct events. These events include fixations, saccades, and optionally postsaccadic oscillations (PSOs) and smooth pursuits. The full algorithm will be available on GitHub (www.github.com/maltelueken/gazeHMM) upon publication.

@@

The HMM estimates non-observable states that have generated the data (CITE). In the context of eye-tracking, each state corresponds to an eye movement event. In gazeHMM, the model can have between two and four states. State one always corresponds to fixations, state two always to saccades, state three to PSOs, and state four to smooth pursuits. Moreover, the HMM is multivariate but conditionally independent on the states. Thus, it has three independent response models for every state describing velocity, acceleration, and angle. The velocity and acceleration response models for all four states use two-parameter gamma distributions (shape and scale parametrization). The angle response model for the fixation state uses a uniform distribution, while the other three states use a two-parameter von-Mises distribution (mean and concentration parametrization) to describe angle. Both the transition model and the initial state model use a multinomial distribution (with the logit and the identity link function, respectively). Note that no covariates are included in the model, only intercepts for the parameters are estimated.

@@

To estimate parameter recovery of the HMM, it will repeatedly generate data with a set of parameters (true parameter values). The same model will then be applied to estimate the parameters from the generated data (estimated parameter values). We will compare the true with the estimated parameter values to assess whether a parameter was recovered by the model. Additionally, we will compare the true states of the HMM with the estimated states to judge how accurate the model recovers the states that generated the data.

## Starting values

The HMM will always start with a uniform distribution to estimate the initial state and state transition probabilities. To generate random starting values for the estimation of shape, scale, and concentration parameters, we will use gamma distributions with the true parameter as the shape parameter and a scale parameter of $\beta = 1$. The gamma distributions ensure that the starting values will be positive. Mean parameters of the von-Mises distribution will always start at their true value.

## Design
### Parameter Variation

The simulation study will be divided into four parts. In the first part, we will vary the parameters of the HMM. For models with $k \in \{2, 3, 4\}$ states, $q \in \{10, 15, 20\}$ parameters will be varied respectively. For each parameter, the HMM will generate 100 data sets with $N = 2500$ samples and the parameter varied in a specified interval in equidistant steps. This will result in $100 \times (10+15+20) = 4500$ recoveries. Only one parameter will be varied at once, the other parameters will be set to their fixed values. We will not manipulate the initial state probabilities because these are usually irrelevant in the context of eye movement classification. For the transition probabilities, we will only vary the probabilities for staying in the same state (diagonals of the tranistion matrix) to reduce the complexity of the simulation. The left probability mass will be split evenly between the probabilites for switching to a different state (per row of the transition matrix). Moreover, we will not modify the mean parameters of the von-Mises distribution: As location parameters, they do not alter the shape of the distribution and they are necessary features for the HMM to distinguish between different states.

Where it was possible, we chose the intervals and fixed values for each parameter based on reported ranges for eye movement events in the literature. Otherwise, intervals and fixed values were chosen according to theoretical reasons and based on insights from data used to develop the model. Table 1 shows the theoretical ranges for each event and Table 2 shows the intervals and fixed values for each parameter in the simulation. Parameters were divided by 10 (compared to the reported ranges) to improve fitting of the gamma distributions. We will set the intervals for shape parameters of the gamma distribution for all events to [1,5] to examine how skewness influences the recovery (shape values above 5 approximate a symmetric distribution). The scale parameters will be set so that the respective distribution will match the ranges reported in the literature. Since the concentration parameters of the von-Mises distribution are the inverse of standard deviations, they will be varied on the inverse scale.

### Sample Size and Noise Variation

In the second part, we will vary the sample size of the generated data and the amount of noise added to it. The model parameters will be set to their fixed values. For models with $k \in \{2, 3, 4\}$ states and sample sizes of $N \in \{500, 2500, 10000\}$, we will generate 100 data sets ($100 \times 3 \times 3 = 900$ recoveries). We chose samples sizes roughly corresponding to small, medium, and large eye-tracking data sets of a single participant and trial. To simulate measurement error, we will add Gaussian white noise with $\mu = 0$ and $\sigma \in [1,5]$ varying between data sets to velocity and acceleration. Values that decrease to zero or below will be set to 0.001. We will add white noise from a von-Mises distribution with $\mu = 0$ and $\kappa \in 1/[0.1,10]$ varying between data sets to angle.

### Variation of Starting Values

In the third part, we will increase the variation in the starting values used for parameter estimation. The model parameters will be set to the fixed values. For the shape, scale, and concentration parameters, we will simultaneously increase the scale parameters of the starting value gamma distributions: For $k \in \{2, 3, 4\}$ states and $\beta \in \{2, 4, 8\}$, 100 data sets with $N = 2500$ samples will be generated each ($100 \times 3 \times 3 = 900$ recoveries).

### Missing data

In the last part, we will set intervals of the generated data to be missing. The model parameters will be set to the fixed values. For $k \in \{2, 3, 4\}$ states and $n_{miss} \in \{1, 3, 5\}$ intervals, 100 data sets with $N = 2500$ samples will be generated ($100 \times 3 \times 3 = 900$ recoveries). The size of the missing data interval $m \in [1,250]$ samples will vary between the data sets.

## Recovery Measures

For each parameter, we will calculate the mean squared error (MSE) between the true and estimated parameter values. Additonally, we will apply a bivariate linear regression  with the estimated parameter values as the dependent and the true parameter values as the independent variable to each parameter that has been varied on an interval. The MSE and the regression coefficients will serve as descriptive measures for parameter recovery. Lower MSE values, regression intercepts closer to zero, and regression slopes closer to one will be interpreted as better recovery.

To assess state recovery, we will compute the proportion of states (for all states taken together, not for each state seperately) that have been classified correctly by the model for each generated data set. Higher proportions will be interpreted as better model accuracy. Overall, we require a median accuracy of 0.9 for the model to be satisfactorily accurate.

@@

*Theoretical ranges of response variables used to generate parameter values*

|Event|Resp. variable|Range|Reference|
|-----|--------------|-----|---------|
|Fixation|Velocity|<30|Larsson et al. (2013)|
|Fixation|Acceleration|<15|-|
|Fixation|Angle|uniform|-|
|Saccade|Velocity|30-500|Larsson et al. (2013)|
|Saccade|Acceleration|10-250|-|
|Saccade|Angle|~0|Pekkanen & Lappi (2017)|
|PSO|Velocity|20-100|Larsson et al. (2013)|
|PSO|Acceleration|10-90|-|
|PSO|Angle|~$\pi$|Pekkanen & Lappi (2017)|
|Smooth pursuit|Velocity|<100|Larsson et al. (2013)|
|Smooth pursuit|Acceleration|<15|-|
|Smooth pursuit|Angle|~0|Larsson et al. (2013)|

*Note*. Units are °/s (velocity), °/s^2^ (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value.

Table 2
